---
title: Data Manipulation (dplyr)
date: '2017-03-13'
slug: data-manipulation-dplyr
categories:
  - R
output:
  blogdown::html_page:
    toc: true  
---

```{r, include=FALSE}
suppressMessages(library(ggplot2)) 
suppressMessages(library(dplyr)) 
suppressMessages(library(babynames)) 
```

Python programmers believe that there should be one -- and preferably only one -- obvious way to do things. In R we love chaos and embrace originality. Therefore, we often have many ways to try.

For data manipulation, we have a solid set of tools in base R like `subset`, `duplicated`, `cbind`, `rbind`, `match` and four ways to use `[` operator (numeric index, negative numeric index, logical index, and character index). For example:

```{r, eval=FALSE, include=FALSE}
var = c(a = 1, b = 2, c = 3, d = 4)
var[1:2]
var[-(3:4)]
var[c(TRUE,TRUE,FALSE,FALSE)]
var[c("a","b")]
```

Here I will talk about package `dplyr` that can be installed as `install.packages(dplyr)` and loaded as `library(dplyr)`. The package gains popularity due to its speed and readable code (`%>%` operator). 



## Dataset babynames

Install and load package `babynames`. Names of male and female babies in US Social Security Administration (from 1880 to 2014). 

Get the info about the dataset with `?babynames` and `View(babynames)`.

## Single table verbs

The are six single table verbs:

* `filter()` - extract cases
* `arrange()` - reorder cases
* `group_by()` - group cases
* `select()` - extract variables
* `mutate()` - create new variables
* `summarise()` - summarise variables / create cases

Download [dData Transformation Cheat Sheet](https://www.rstudio.com/resources/cheatsheets/) to see other `dplyr `data transformations.

### filter

Extract rows that meet logical criteria, see `?filter` for help. Also, see `?Comparison` and `?base::Logic` for logical operators in R.

#### Challenge 1

Try to plot a proportion of the name Wendy in US population.

```{r, echo=FALSE}
wendy <- filter(babynames, name == "Wendy", sex == "F")
ggplot(wendy, aes(x=year, y=prop)) +
  geom_line() +
  labs(title = "Wendy")
```

### arrange

Order rows from smallest to largest values. See `?arrange` for help. 

#### Challenge 2

What was the most popular name in a single year? In what year was Wendy the most popular?

```{r, eval=FALSE, include=FALSE}
arrange(babynames, -prop) # John in 1880
wendy <- filter(babynames, name == "Wendy", sex == "F")
# instead of -prop, you can also use desc(prop)
arrange(wendy, desc(prop)) #1967
```

### select 

Extract columns by name. See `?select` and *select helpers* in the cheat sheet.

### mutate

Create new columns, see `?mutate`.

#### Challenge 3

Create a new table `babynames2` with a new percent column `perc = 100*prop`, then drop the column `prop`.

```{r, echo=FALSE}
tmp = mutate(babynames, perc = 100 * prop)
babynames2 = select(tmp, -prop)
babynames2
```

### summarise

Compute table of summaries.

#### Challenge 4

* Create a summary that displays the min, mean, and max prop for Wendy. 
* Determine how many unique names appear in the data set. (see cheat sheet or help for `n_distinct`).
* How many times a single names was given to $1\%$ or more of the boys or girls?

```{r, include=FALSE}
wendy = filter(babynames, name == "Wendy")
summarize(wendy, min(prop), mean(prop), max(prop))

summarize(babynames, n = n_distinct(name))

babynames2 <- filter(babynames, prop > 0.01)
summarise(babynames2, nn = n())
```

## Operator %>%

A little bunny hopped though the forrest, scooped up the field mouse and bopped on his head. It could be expressed as follows.

```{r, eval=FALSE}
foo_foo <- little_bunny()
foo_foo2 <- hop_through(foo_foo, forest)
foo_foo3 <- scoop_up(foo_foo2, field_mouse)
bop_on(foo_foo3, head)
```

Or the equivalently as...

```{r, eval=FALSE}
bop_on(
  scoop_up(
    hop_through(little_bunny(), forest),
    field_mouse
  ),
head)
```

The operator `%>%` enables us to sequentially describe the procedure without creating extra variables.

```{r, eval=FALSE}
little_bunny() %>%
  hop_through(forest) %>%
  scoop_up(field_mouse) %>%
  bop_on(head)
```

## Grouping cases

## Joining data sets