---
title: Tidy Data (tidy, purrr)
date: '2017-03-14'
slug: tidy-data-tidy-purrr
categories:
  - R
output:
  blogdown::html_page:
    toc: true
---

```{r, include=FALSE}
suppressMessages(library(tidyverse)) 
suppressMessages(library(babynames))
data(iris)
```

We have learned basic data manipulation with `dplyr`. Now it is time to learn reshaping of tabular data with `tidyr` and looping through elements with `purrr`

## dplyr

* `spread` - move values into column names
* `gather` - move column names into values
* `separate` - separate variables that share a column
* `unite` - unite a variable that is split across several columns

### spread

Imagine you want to calculate a ratio of number of males to females in `babynames` dataset. You can start as 

```{r, eval=FALSE}
babynames %>%
  group_by(year, sex) %>%
  summarise(n = sum(n))
```

And then you need something that from `n` generates two columns `males` and `females`. 

```{r}
babynames %>%
  group_by(year, sex) %>%
  summarise(n = sum(n)) %>%
  spread(sex, n) %>%
  head()
```

#### Challenge 1:

Modify the code above to plot the percent of childern that are male in `babynames` dataset:

```{r, echo=FALSE}
babynames %>%
  group_by(year, sex) %>%
  summarise(n = sum(n)) %>%
  spread(sex, n) %>%
  mutate(percent_male = M / (M + F) * 100) %>%
  ggplot(aes(year, percent_male)) + geom_line()
```

### gather

`gather` is inverse function to `spread`, its collapses multiple columns into two columns, see `?gather`.

#### Challenge 2:

For the wide transformation of `babynames`, use `gather` to transform the dataset back into the long format with three columns. 

```{r}
wide <- babynames %>%
  group_by(year, sex) %>%
  summarise(n = sum(n)) %>%
  spread(sex, n)
```

```{r, eval=FALSE, include=FALSE}
# solution
long <- wide %>%
  gather("sex", "n", 2:3)
```

### separate and unite

`separate` splits a column by dividing values at a specific character. `unite` unites columns into single column by combining cells.

### iris dataset 

Famous Fisher's dataset with measurements of sepal / petal length and width for *Iris setosa, versicolor*, and *virginica*.

Let us start with some data inspection. What is the distribution of `Sepal.Length` for each specie?

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_boxplot() +
  geom_jitter(width=0.2)
```

Instead of boxplots and data points, we can plot violins and means.

```{r}
stats <- iris %>%
  group_by(Species) %>%
  summarize(average = mean(Sepal.Length),
            se = sd(Sepal.Length) / sqrt(n())) %>%
  ungroup()

ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  geom_violin() +
  geom_pointrange(data= stats, aes(x = Species, y = average, 
                                    ymin = average-se, ymax = average+se), size=0.5)
```

Or if somebody insist - barplots and error bars.

```{r}
ggplot(stats, aes(x = Species, y = average, fill = Species)) + 
  geom_bar(stat="identity") +
  geom_errorbar(data= stats, aes(ymin = average-se, ymax = average+se), width=0.5) +
  ylab("Sepal.Length")
```

