---
title: Grammar of Graphics (ggplot2) [Part 2]
date: '2017-03-07'
slug: ggplot2-part-2
categories:
  - R
output:
  blogdown::html_page:
    toc: true
---

```{r, include=FALSE}
suppressMessages(library(ggplot2)) 
data(diamonds)
```

## ggplot2 recap

* Input data into `ggplot`
* Add visualization layers with `geom_` functions
* Map *variables* to aesthetics with `aes()`
* Plot summaries with `stat=`
* Handle collisions with `position=`
* Change coordinate systems with a `coord_` function
* Facet the plot with a `facet_` function

#### Challenge 1:

Try to make the following plot (density plot of `price` in dependence on `cut`) and save it into variable `p1`.

```{r, echo=FALSE}
p1 <- ggplot(diamonds, aes(x = price, color = cut)) +
  geom_density()
p1
```

## labels

A function to add labels to a plot is `labs`. You can add a title, subtitle and caption as follow.

```{r}
p1 + labs(title = "Poorly cut diamonds appear to fetch unexpectedly high prices",
      subtitle = "The lowest cut rating is associated with the highest mode price",
      caption = "Data by Hadley Wickham")
```

Similarly, you can change axis labels and legend.

```{r}
p1 + labs(x = "Price", y = "Density", color = "Cut Rating")
```

## themes

Themes change appearance of non-data elements.

First, you change theme manually using `theme` function. See the list of what you can control in [the documentation](http://docs.ggplot2.org/current/theme.html). For example, you can take the previous plot and remove the legend:

```{r}
p1 + theme(legend.position = "none")
```


Second, you can use predefined themes. Take the `p1` plot and try to add the following themes:

* `p1 + theme_bw()`
* `p1 + theme_classic()`
* `p1 + theme_dark()`
* `p1 + theme_light()`
* `p1 + theme_linedraw()`
* `p1 + theme_minimal()`
* `p1 + theme_void()`

Many developers come with their own ggplot2 themes. For example with `library(ggthemes)` you can try `p1 + theme_wsj()` to style your plot like Wall Street Journal, with `library(xkcd)` you can style your plot as [xkcd comics](https://xkcd.com/1798/) and Bob Rudis creates his [hrbrthemes](https://github.com/hrbrmstr/hrbrthemes) for his style.

```{r}
# devtools package can be installed as install.packages("devtools")
# hrbrthemes package can be then installed as devtools::install_github("hrbrmstr/hrbrthemes")
library(hrbrthemes)
p1 + theme_ipsum()   # or theme_ipsum_rc() on MacOS  
```

## scales

You can add a `scale_` function for each aesthetics.

Let us plot the dependence between price and weight (`carat`):

```{r}
s <- ggplot(diamonds, aes(carat, price)) +
   geom_point(aes(color = cut))
s
```

#### Challenge 2:

Try to apply various scales to the graph `s`:

* for color, try `scale_color_grey` and `scale_color_brewer(palette = "Set1")`
* for axis, try `scale_x_log10()` and `scale_y_log10()`

#### Challenge 3:

If we now color previous graph by `price` instead of `cut`:

```{r}
s1 <- ggplot(diamonds, aes(carat, price)) +
   geom_point(aes(color = price))
```

Try to apply various continuous color scales.

* from `library(viridis)`: `scale_color_viridis()` and `scale_color_viridis(option = "A")`
* `scale_color_distiller(palette = "Spectral")`
* `scale_color_brewer(palette = "Purples")`

#### Challenge 4:

Now experiment with labels, themes, and scales to make a clear graph how a price of diamond depends on its weight. Perhaps this one, or any other you like.

```{r, echo=FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() +
  geom_smooth(aes(color = cut), se = FALSE) +
  labs(title = "Ideal cut diamonds command the best price for ever carat size",
       subtitle = "Lines show GAM estimate of mean values for each level of cut",
       caption = "Data provided by Hadley Wickham",
       x = "Log Carat Size",
       y = "Log Price Size",
       color = "Cut Rating") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(palette = "Greens") +
  theme_light()
```

## saving graphs

There is more than one way to save the graph:

* In RStudio you can use `Export` tab in `Plots` panel
* In `ggplot2` package you can use a `ggsave` function (see `?ggsave` for help and examples)
* You can `pdf`, `png` and `dev.off` as in base graphics 

#### Challenge 5:

Save the plot generated in Challenge 4 into PDF file.

```{r, eval=FALSE, include=FALSE}
ggsave("my-plot.pdf", width = 6, height = 6)
```

## summary 

`ggplot2` in a combination with custom themes are a fast route to creation of (almost) publication-quality graphics.

#### Sources:

* Data Visualization with ggplot2 (rstudio::conf, Tidyverse workshop)
* [ggplot2 documentation](http://ggplot2.org/)

