---
title: R Base Graphics
date: '2017-02-28'
slug: r-base-graphics
categories:
  - R
output:
  blogdown::html_page:
    toc: true
---

One of the strengths of R is its ability to produce a wide range of plots. Multiple plotting libraries exist and even the base R graphics is powerful and can be mastered in a few hours.

## X-Y Plotting

Basic plot call takes the following form `plot(x=, y=)`, for example

```{r}
height <- c(123, 145, 150, 167, 176)
weight <- c(40, 51, 55, 63, 64)

plot(height, weight)
```

Use `main` to set the title, `xlab` and `ylab` for x- and y- labels, `xlim` and `ylim` for x- and y- limits, `type` to set a type of the graph ("p"=points, "l"=lines, "b"=both, "n"=none) and `col` to choose colors.

#### Challenge 1:

Modify the code above to get points connected by a line

```{r, echo=FALSE}
plot(height, weight, type="l")
```

Or to see both points and a line like on the plot below.

```{r, echo=FALSE}
plot(height, weight, type="b")
```

To the existing plot, you can add lines (`lines`, `abline`), points (`points`), legend (`legend`), text (`text`) and other objects.

## Gapminder dataset

Gapminder is a an excerpt of the Gapminder data on life expectancy, GDP per capita, and population by country. It is distributed as R package. The following code loads `gapminder` (if missing, installs it) and prints the first few lines:

```{r}
# check if gapminder can be loaded, if not install it
if (!require("gapminder")) {
  install.packages("gapminder")
  library("gapminder")
}

data(gapminder)
head(gapminder)
```

Type `?gapminder` into your console to get info about the dataset and try to answer the following questions.

#### Challenge 2

How many rows and columns are in gapminder? What was the population of Czech Republic in 1952? Which country has the highest GDP per capita in 2007?

```{r, include=FALSE}
# number of rows and columns 
nrow(gapminder)
ncol(gapminder)
dim(gapminder)

# life expectancy in 1952 Czechia
subset(gapminder, country=="Czech Republic" & year == 1952)

# highest GDP per capita in 2007
year2007 <- subset(gapminder, year == 2007)
max.idx <- which.max(year2007$gdpPercap)
year2007[max.idx,]
```

#### Challenge 3

For 2007, make a scatter-plot of life expectancy vs. GDP per capita and color each state by its continent. 

```{r, echo=FALSE}
year2007 <- subset(gapminder, year == 2007)
gdpPercap2007 <- year2007$gdpPercap
lifeExp2007 <- year2007$lifeExp
# use as.numeric to convert factor to integer (+1 to avoid black)
col <- as.numeric(year2007$continent) + 1

plot(x=gdpPercap2007, y=lifeExp2007, col=col)
```

Now try to style the plot a bit by adding the title, labels and a legend.

```{r, echo=FALSE}
plot(x=gdpPercap2007, y=lifeExp2007, col=col, main="2007",
     xlab = "GDP per capita [US dollars]", ylab = "life expectancy at birth [years]", pch=19)
legend("bottomright", levels(year2007$continent), col=2:6, pch=19)
```

## Histogram

Histograms are plotted with `hist` function (see `?hist` for help and examples). 

#### Challenge 4

Plot a histogram of life expectancy in 2007 over all countries:
```{r, echo=FALSE}
year2007 <- subset(gapminder, year == 2007)
lifeExp2007 <- year2007$lifeExp

hist(x = lifeExp2007)
```

## Barplot

Barplots are plotted with `barplot` function. 

#### Challenge 5

Visualize how the population of the Czech Republic grows by a line and a barplot. Which plot do you prefer?

```{r, echo=FALSE}
czech <- subset(gapminder, country == "Czech Republic")

plot(x = czech$year, y = czech$pop / 1e6, type="l", las=2)
barplot(czech$pop / 1e6, ylab="population [in millions]")
```

## More than one plot in a window

With a graphical parameter `mfrow` we can get two or more plots into one window (or file). For example, see the comparison of the histogram to a density plot below.

```{r}
czech <- subset(gapminder, country == "Czech Republic")

par(mfrow = c(1,2)) # one row, two columns
hist(czech$gdpPercap, main="Histogram")
plot(density(czech$gdpPercap), main="Density")
```

## Boxplot

Boxplots can be used to compare data from several groups. 

#### Challenge 6

For 2007, compare life expectancy between the continents.

```{r}
year2007 <- subset(gapminder, year == 2007)

# las = 2 plots text perpendicularly to axis
boxplot(lifeExp ~ continent, data=year2007, las=2)
```

## Saving the plots

See the help and examples for `pdf`, `png`, `dev.off` and `dev.copy` functions.

#### Challenge 7

Save the last plot into PDF and PNG files.

```{r, eval=FALSE, include=FALSE}
# png
png("boxplot.png", width = 500, height = 500, res = 72)
  boxplot(lifeExp ~ continent, data=year2007, las=2)
dev.off()

# pdf
pdf("boxplot.pdf", width = 6, height = 6)
  boxplot(lifeExp ~ continent, data=year2007, las=2)
dev.off()
```

Are the plots same or different? Why?

## Summary

You can do almost anything with base graphics but some common actions are not straightforward (nice colors, faceting, legends, error bars). Complex graphs can be time consuming and their code hard to read. 

R Base graphics is great for a quick exploratory analysis but for a publication level graphics, I prefer [ggplot2](http://ggplot2.org/). More about it next time.

#### Sources:

* [http://rpubs.com/SusanEJohnston/7953](http://rpubs.com/SusanEJohnston/7953)
* [http://www.statmethods.net/graphs/scatterplot.html](http://www.statmethods.net/graphs/scatterplot.html)
* [http://www.statmethods.net/advgraphs/parameters.html](http://www.statmethods.net/advgraphs/parameters.html)