---
title: Regular Expressions and Text Processing
date: '2017-03-20'
slug: regular-expressions-and-text-processing
categories:
  - R
output:
  blogdown::html_page:
    toc: true  
---

R is not known to be a favorite tool for text processing. However, it contains basic functions to concatenate and split strings. 

## text manipulation

### base

* `paste`, `paste0` - concatenate multiple strings
* `strplit` - split one string into multiple strings
* `substr` - get a substring

#### Challenge 1:

If `a = "R is awesome"` and `b = "but sometimes sucks"` then 

* Combine those strings together
* Split the first one by " " character 
* Get "sometimes" from the second one using `substr`

```{r, echo=FALSE}
a = "R is awesome"
b = "but sometimes sucks"
paste(a, b)
strsplit(a, " ")[[1]]
substr(b, 5, 13)
```

### stringr

I have very limited knowledge of `stringr` but it is considered to be more friendly way to do string manipulation in R. It is now a part of `tidyverse` collection.

See https://github.com/tidyverse/stringr

#### Challenge 2:

If you like `stringr`, try to do Challenge 1 with `stringr` functions.

```{r, echo=FALSE}
library(stringr)
a = "R is awesome"
b = "but sometimes sucks"
str_c(a, b, sep = " ")
str_split(a, " ")[[1]]
str_sub(b, 5, 13)
```

## regex

Regular expressions (regex) allow you to search for pattern inside text. The very basic idea is demonstrated here:

http://www.dummies.com/programming/r/how-to-use-regular-expressions-in-r/

In R we can use `grep` function to search for patterns, for example

```{r}
strings <- c("a", "ab", "acb", "accb", "acccb", "accccb")
grep("ac*b", strings, value = TRUE)
```

The patter "ac*b" stands for first "a", then any number of "c", then "b".

#### Challenge 3:

In `library(gapminder)` use `gapminder` dataset, column `country` to search for all countries with `ee`.

```{r, echo=FALSE}
library(gapminder)
countries <- unique(gapminder$country)
grep("ee", countries, value=TRUE)
```

### operators

* `.`: matches a single character
* `[...]`: matches any one of the characters inside the square brackets
* `[^...]`: similar to  `[...]`, but matches any characters except those inside the square brackets.
* `\`: suppress the special meaning of metacharacters in regular expression, i.e.  $ * + . ? [ ] ^ { } | ( ) \
* `|`: an “or” operator, matches patterns on either side of the |.
* `^` and `$`: beginning and end of the string

#### Challenge 4:

What will be the output of the code below (first guess, then try).

```{r, eval=FALSE}
strings <- c("^ab", "ab", "abc", "abd", "abe", "ab 12")
grep("ab.", strings, value = TRUE)
grep("ab[c-e]", strings, value = TRUE)
grep("ab[^c]", strings, value = TRUE)
grep("^ab", strings, value = TRUE)
grep("\\^ab", strings, value = TRUE)
grep("abc|abd", strings, value = TRUE)
```

### substitutions

With `sub` and `gsub` functions you can make one or all string substitutions. 

```{r}
sub("a", "b", "balalajka")
gsub("a", "b", "balalajka")
```

There is one additional regex operator for substitutions:

* `(...)`: grouping in regular expressions. This allows you to retrieve the bits that matched various parts of your regular expression so you can alter them or use them for building up a new string. Each group can than be refer using \\N, with N being the No. of (...) used. This is called backreference.

For example, to add dash before the number in sample ids.

```{r}
ids <- c("DO01", "DO02", "DO03", "DO04", "DO05")
sub("DO([0-9]*)", "DO-\\1", ids)
```


#### Challenge 5:

In `gapminder`, find all countries that end with "land", and replace "land" with "LAND" using backreference. 

```{r}
countries <- unique(gapminder$country)
lands <- grep("land$", countries, value = TRUE)
sub("(.*)land", "\\1LAND", lands)
```

## summary

#### Sources:

* [STAT545 Regular Expression in R](http://stat545.com/block022_regular-expression.html)
* [R help](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html)